
---
title: "Fixed Income Risk Analysis – Case Study"
author: "Noluvuyo Yumata"
output:
  slidy_presentation: default
  html_document: null
  pdf_document:
    toc: true
  powerpoint_presentation:
    reference_doc: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,   # hide code by default
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 7,       # fits PowerPoint 16:9 slides
  fig.height = 4.5
)



library(gt)
library(dplyr)
library(readxl)
library(ggplot2)
library(stringr)
library(knitr)  
library(scales)
library(tidyverse)
library(janitor)
library(lubridate)
library(patchwork)
library(openxlsx)
library(stringr)
library(flextable)
library(tidyr)
library(quantmod)
library(tidyquant)
library(kableExtra)
library(ggrepel)

file <- "FI Risk Analyst Case Study.xlsx"
```

```{r, include= FALSE}
#---------------------------------------------- DATA EXPLORATION ---------------------------------------------------------------
# Load snapshots -----------------------------------------------------
mar <- read_excel(file, sheet = "FI Data (31 March 2024)")
jun <- read_excel(file, sheet = "FI Data (30 June 2024)")

# Load risk‑metric history ------------------------------------------
risk_ts <- read_excel(file, sheet = "Risk Metrics (Time Series)") %>%
  clean_names() %>%
  mutate(reference_date = as.Date(reference_date))

# Combine & clean snapshot data -------------------------------------
snaps_raw <- bind_rows(
  mar %>% mutate(snapshot = as_date("2024‑03‑31")),
  jun %>% mutate(snapshot = as_date("2024‑06‑30"))
) %>%
  clean_names() %>%
  mutate(
    dirty_price = readr::parse_number(dirty_price),
    across(matches("(percent|pct|%)$", ignore.case = TRUE), as.numeric)
  )

summary(snaps_raw)
```

```{r, include = FALSE }


#------------------------  Issuer Classification --------------------------------------------------------------------------------------------------------------------------

classify_issuer <- function(name) {
  name <- str_to_upper(as.character(name)) #Comparison regardless of text format
  case_when(
    str_detect(name, "NEDBANK") ~ "Nedbank",
    str_detect(name, "ABSA") ~ "ABSA",
    str_detect(name, "STANDARD BANK") ~ "Standard Bank",
    str_detect(name, "FIRSTRAND") ~ "FirstRand",
    str_detect(name, "INVESTEC BANK") ~ "Investec Bank",
    str_detect(name, "CORBON|NINETY ONE") ~ "Ninety One",
    str_detect(name, "SOUTH AFRICA, REPUBLIC") | str_detect(name, "SOUTH AFRICA\\s+R\\d{3,4}")~ "SA Govi",
    str_detect(name, "NAMIBIA, REPUBLIC") ~ "Namibia Govi",
    str_detect(name, "LIBERTY") ~ "Liberty",
    str_detect(name, "OLD MUTUAL") ~ "Old Mutual",
    str_detect(name, "MOMENTUM") ~ "Momentum",
    str_detect(name, "SANLAM") ~ "Sanlam",
    str_detect(name, "SANTAM") ~ "Santam",
    str_detect(name, "REDEFINE") ~ "Redefine",
    str_detect(name, "GROWTHPOINT") ~ "Growthpoint",
    str_detect(name, "NORTHAM") ~ "Northam",
    str_detect(name, "CAPE TOWN") ~ "City of CPT",
    str_detect(name, "TSHWANE") ~ "City of Tshwane",
    str_detect(name, "AIRPORTS CO") ~ "ACSA",
    str_detect(name, "SOUTH AFRICAN NATIONAL ROADS") ~ "Sanral",
    str_detect(name, "TRANSNET") ~ "Transnet",
    str_detect(name, "ESKOM") ~ "Eskom",
    str_detect(name, "TELKOM") ~ "Telkom",
    str_detect(name, "MOBILE TELEPHONE NETWORKS") ~ "MTN",
    str_detect(name, "MOMENTUM") ~ "Momentum",
    str_detect(name, "RAND WATER") ~ "Rand water",
    str_detect(name, "TOYOTA") ~ "Toyota",
    str_detect(name, "THEKWINI") ~ "Thekwini Fund",
    str_detect(name, "NQABA FINANCE") ~ "Nqaba Finance",
    str_detect(name, "GREENHOUSE FUNDING") ~ "Greenhouse Fund",
    str_detect(name, "SOUTH AFRICAN RAND") ~ "Cash",
    TRUE ~ "Other / Unmapped"
  )
}
snaps_raw$issuer <- sapply(snaps_raw$asset_name, classify_issuer)

```

```{r, incude = FALSE}
#--------------------------------------------- Asset Class classification ----------------------------------------------------------------------------------------


classify_class <- function(df){
  df %>%
    mutate(
      asset_type = case_when(

        # --- look at coupon_type OR asset_name --------------------
        asset_id == "ZAR" | str_detect(asset_name, regex("ZAR|CASH|South African Rand", TRUE))
                      ~ "Cash",

        str_detect(asset_name, regex("\\TREASURY |SOUTH AFRICA, REPUBLIC | \\s+R\\d{3,4}", TRUE))
                      ~ "SAGBs",
        
        str_detect(asset_name, regex("BANK| ABSA| STANDARD BANK | FIRSTRAND |CAPITEC|NEDBANK", TRUE))
                      ~ "SA Banks",

        str_detect(asset_name, regex("THEKWINI|NQABA|FUND|SECURIT|SPV|CMBS|RMBS", TRUE))
                      ~ "Securitisation",
        
        str_detect(asset_name, regex("NAMIBIA", TRUE))
                      ~ "Offshore govi",
   TRUE  ~ "Corporate")
     
    )
}

class_classif <- classify_class(snaps_raw)

structured <- c( "SSW240606",
"SSW240607",
"SSW240608",
"SSW240609",
"SSW231206",
"SSW240201"
 )

future <- c("RDQ24",
            "ZBK24",
            "REQ24",
            "RFQ24",
            "RDQ24",
            "REQ24",
            "RFQ24",
            "ZBQ24",
            "ZNQ24",
            "ZPQ24",
            "ZSQ24",
            "ZTQ24",
           "RDK24",
            "REK24",
            "RFK24",
            "ZBK24",
            "ZNK24",
            "ZPK24",
            "ZSK24",
            "ZTK24"

            )

repos <- c("006XNO",
           "02WN0S")


structured <- c("SSW240606",
                "SSW240607",
                "SSW240608",
              "SSW240609"
)

zero_cpn <- c("CORBON")


classify_asset_type<- function(df){
  df %>%
    mutate(
      asset_type = case_when(

        # --- look at coupon_type OR asset_name --------------------
        asset_id == "ZAR" | str_detect(asset_name, regex("ZAR|CASH|South African Rand", TRUE))
                      ~ "Cash",

        str_detect(asset_name, regex("\\TREASURY |SOUTH AFRICA, REPUBLIC | \\s+R\\d{3,4}", TRUE))
                      ~ "SAGBs",

        asset_id =="NSW231203" | str_detect(asset_name, regex("JIBAR |JIB|FRN|FLOAT", ignore_case = TRUE)) 
                      ~ "Swaps",
        

        str_detect(asset_name, regex("THEKWINI|NQABA|FUND|SECURIT|SPV|CMBS|RMBS", TRUE))
                      ~ "Securitisation",
        
        asset_id == "CORBON" ~ "Zero Coupon" ,
        
        asset_id %in% structured ~ "Structured" ,
        
        asset_id %in% future ~ "Futures" ,
        
        asset_id %in% repos ~ "Repos",

        !is.na(asset_name)| asset_id =="ZAG000138470" ~ "Corporate",

        TRUE             ~ "Other"      # default bucket
      )
    )
}

type_classif <- classify_asset_type(class_classif)

```


```{r, include = FALSE }
#_____________________________________________ ANOMALITY DETECTION & FLAGGING --------------------------------------------------------------------------------------
snap_flagged <- function(x) {
  x %>%
  mutate(price_zero   = price <= 0,
         price_na     = is.na(price),
         dur_extreme  = active_effective_duration_mac > 100,
         neg_holdings = holdings < 0,
         high_risk     = weight_percent <=0 & abs(active_total_risk) > 50, # update these thereshold if needed
         flag_any = price_zero|price_na| dur_extreme|neg_holdings| high_risk
    ) %>%
    filter(flag_any)
}

flag_obs  <- snap_flagged(type_classif)

# ──--------------- Side‑by‑side summary table ───────────────────────────────────────────
summary_tbl <- flag_obs %>%
  pivot_longer(cols = c(price_zero,price_na, dur_extreme,neg_holdings, high_risk),
               names_to = "Flag", values_to = "Triggered") %>%
  filter(Triggered) %>%
  group_by(snapshot, Flag) %>%
  summarise(Count = n(), .groups = "drop") %>%
  pivot_wider(names_from = snapshot, values_from = Count, values_fill = 0) %>%
  arrange(Flag) 

#print(summary_tbl)


#-------------------------- Review & Automate the flagged observations

#---- This anomalities are saved straight to the folder

wb <- createWorkbook()
addWorksheet(wb, "flag_obs");  writeData(wb, "flag_obs", flag_obs)
saveWorkbook(wb, "Flagged_abnomalies.xlsx", overwrite = TRUE)


```


```{r, include=FALSE}

#----------------------------------------- PRICE OVERRIDES - TEMP PLACE HOLDERS--------------------------------------------------------------------------------------
# approc Price with par and Corporate with par 100

pric_adj <- type_classif %>%
  mutate(price = if_else(asset_id %in% c("ZAR"), 1, price),
         price = if_else(asset_type %in% c("Corporate","SA Banks"), 100, price))
  

#------------------------------------ HEDGE AdJuSTMENTS ------------------------------------------------------------------------------------------------------------

hedge_df <- pric_adj %>%
  mutate(hedging = if_else(asset_type %in% c("Future","FRN"), "hedge","non hedge")) 
  
#----------------------------------CLEANED PORTFOLIO ---------------------------------------------------------------------------------------------------------------

#For negative positions (which with no material impact on NAV)- kept them position but assigned zero holdings 
snap_clean <- hedge_df %>%
 mutate(holdings = ifelse(holdings<0,0,holdings)) %>% # this is to neutralise the positions
mutate(mv = price*holdings) %>%
  mutate(holdings_weight=holdings/sum(holdings))


#see if the adjustments
#summary(snap_clean) 
```


---

# **Global Macro Enviroment**

* High-Interest Rate volatility and slow easing to Disinflation
*	1Q24- Due to sticky inflation, Economic Instability & Uncertanity around geopolitical 
* 2Q24- Advance economies saw inflation gradually decoding leading to market to factor first rate cuts late in 2024/2025
* Euro zone being the first to cut rates in March 2024 by 25bp Due to their CPI falling below their target sat of 2%. 
* Swiss Low Inflation, low- growth mix allowed it to set the tone for Global rate cut cycle.
*	US CPI has fallen from a high of 9.1 % (Jun-22) to the mid-2 % zone; core-PCE is now just below 2.5%.

---

# **Local Macro Enviroment**

* Highly Volatile Market(1Q24) and improving risk sentiments, boosting Investor Confidence (2Q24) 
* *Inflation & MPC Policy*	CPI ~ 5.2% to 5.6% (above SARB mid target) and tilted towards 4%.
* *Economic Activity*	 Weak to Stabilising GDP growth & improving Business sentiment.
  1Q24 GDP(-0,3%) to 2Q24 (0.4) from mining, manufacturing,  Transport.
  *Load shedding eased* Improving business and production constraints
* *Structural and Fiscal factor* Infrastructure bottleneck, high Unemployment, High Govi debt (74% of GDP) and 2024 Election results.
  GNU boosted Investor confidence, optimism towards accelerated reforms and embedded policy stability.
* *SARB held rates steady* at 8.25% indicating repo-rate cuts late in 2Q24 despite signs of improving Macro conditions (both Globally and locally)

---


# Portfolio Headline - Jun 24

```{r headline }
headline_stat <- snap_clean %>%                  # ← your original aggregation
  group_by(snapshot) %>%
  summarise(
    NAV           = sum(holdings),
    total_mv      = sum(holdings * price),
    port_duration = weighted.mean(active_effective_duration_mac,
                                  w = weight_percent),
    spread_dur    = weighted.mean(active_spread_duration,
                                  w = weight_percent),
    n_positions   = n(),
    .groups = "drop"
  )


june_tbl <- headline_stat %>%                          # already created above
  filter(snapshot == as.Date("2024-06-30")) %>%
  select(-snapshot) %>%
  mutate(across(c(NAV, total_mv), ~ .x / 1e9)) %>%     # R-billions
  pivot_longer(everything(),
               names_to  = "Metric",
               values_to = "Value") %>% 
  mutate(                                              # pretty labels
    Metric = recode(Metric,
      NAV           = "NAV (R bn)",
      total_mv      = "Total MV (R bn)",
      port_duration = "Eff. MacDur (yrs)",
      spread_dur    = "Spread Dur (yrs)",
      n_positions   = "No. Positions"
    )
  )

# ── Presentation table with gt ──────────────────────────────────────
june_tbl %>% 
  gt(rowname_col = "Metric") %>%      # metrics become row labels

  ## Header ----
  tab_header(
    title    = md("**Portfolio Headline Metrics**"),
    subtitle = md("As at&nbsp;30&nbsp;June&nbsp;2024")
  ) %>% 

  ## Number formatting ----
  fmt_number(
    rows = Metric %in% c("NAV (R bn)", "Total MV (R bn)"),
    decimals = 1
  ) %>% 
  fmt_number(
    rows = Metric %in% c("Eff. MacDur (yrs)", "Spread Dur (yrs)"),
    decimals = 2
  ) %>% 
  fmt_integer(
    rows = Metric == "No. Positions",
    sep_mark = ","
  ) %>% 

  ## Styling ----
  cols_label(Value = "") %>%          # hide column header
  cols_align(align = "right") %>% 
  cols_width(everything() ~ px(180)) %>% 
  tab_options(
    table.font.names        = "Verdana, Arial, sans-serif",
    data_row.padding        = px(4),
    table.border.top.style  = "hidden",
    table.border.bottom.style = "hidden"
  ) %>% 
  # subtle banding
  tab_style(
    style = cell_fill(color = "#f7f7f7"),
    locations = cells_body(rows = seq(2, nrow(june_tbl), 2))
  )

```

* The porfolio has a nominal value of R4,6% bn and mainly concentrated around long dated govies with strategic tits using credit ( NinetyOne Bond at ~60%, ASW ( 25%) then GOV futures. It has a mac duration -0.76 and spread duration of -0.65  Suggest a strategic tilt around short-dated instruments & strong credit using of derivates instruments including swaps and futures. 


---

# Asset Mix (% Holdings)


```{r asset_nominal include}

june  <- snap_clean %>%
         filter(snapshot == as.Date("2024-06-30"))

# ── 2  Presentation theme with extra title spacing ────────
theme_present <- function() {
  theme_minimal(base_size = 12, base_family = "Arial") %+replace%
    theme(
      plot.title      = element_text(size = 14, face = "bold",
                                     hjust = 0.5, margin = margin(b = 15)),
      plot.subtitle   = element_text(size = 11,  hjust = 0.5,
                                     margin = margin(b = 6)),
      axis.title.y    = element_blank(),
      axis.title.x    = element_text(margin = margin(t = 6)),
      axis.text       = element_text(color = "grey30"),
      panel.grid.major.y = element_blank(),
      legend.position = "none",
      plot.margin     = margin(10, 20, 10, 10)  # TLBR
    )
}

# Corporate colour (same for all bars)
corp_blue <- "#2C7FB8"

# Helper to add % labels above bars ------------------------
add_pct_labels <- function(df, pct_col) {
  geom_text(aes(label = percent({{ pct_col }}, accuracy = 0.1)),
            hjust = -0.15, size = 3.3)
}

# ── 3-A  Asset-type allocation (% holdings) ─────────────────────
asset_alloc <- june %>%
  group_by(asset_type) %>%
  summarise(holdings = sum(holdings, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = holdings / sum(holdings))

asset_mix_nom <- ggplot(asset_alloc,
       aes(x = fct_reorder(asset_type, pct), y = pct)) +
  geom_col(width = 0.7, fill = corp_blue) +
  add_pct_labels(asset_alloc, pct) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(asset_alloc$pct) * 1.15)) +
  labs(title = "Asset Mix (% Nominal)",
       subtitle = "30 Jun 2024  . % of portfolio nominal",
       y = "% of nominal") +
  theme_present()

print(asset_mix_nom)
```
---

# Asset Mix (% MV)

```{r asset_m }

june  <- snap_clean %>%
         filter(snapshot == as.Date("2024-06-30"))

# ── 2  Presentation theme with extra title spacing ────────
theme_present <- function() {
  theme_minimal(base_size = 12, base_family = "Arial") %+replace%
    theme(
      plot.title      = element_text(size = 14, face = "bold",
                                     hjust = 0.5, margin = margin(b = 15)),
      plot.subtitle   = element_text(size = 11,  hjust = 0.5,
                                     margin = margin(b = 6)),
      axis.title.y    = element_blank(),
      axis.title.x    = element_text(margin = margin(t = 6)),
      axis.text       = element_text(color = "grey30"),
      panel.grid.major.y = element_blank(),
      legend.position = "none",
      plot.margin     = margin(10, 20, 10, 10)  # TLBR
    )
}

# Corporate colour (same for all bars)
corp_blue <- "#2C7FB8"

# Helper to add % labels above bars ------------------------
add_pct_labels <- function(df, pct_col) {
  geom_text(aes(label = percent({{ pct_col }}, accuracy = 0.1)),
            hjust = -0.15, size = 3.3)
}

# ── 3-A  Asset-type allocation (% holdings) ─────────────────────
asset_alloc_mv <- june %>%
  group_by(asset_type) %>%
  summarise(mv = sum(mv, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = mv / sum(mv))

asset_mix_mv <- ggplot(asset_alloc,
       aes(x = fct_reorder(asset_type, pct), y = pct)) +
  geom_col(width = 0.7, fill = corp_blue) +
  add_pct_labels(asset_alloc, pct) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(asset_alloc$pct) * 1.15)) +
  labs(title = "Asset Mix (% Market value)",
       subtitle = "30 Jun 2024  .  % of portfolio market value",
       y = "% of mv") +
  theme_present()

asset_mix_mv 

#----------------------------------------- BOTH ASSET MIX BY NOMINAL AND MV___________________________________
```

---

# Issuer(%Holdings)

```{r }
# ── 3-C  Issuer allocation (% holdings) ─────────────────────────
issuer_alloc <- june %>%
  group_by(issuer,asset_type) %>%         # change if your column differs
  summarise(holdings = sum(holdings, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = holdings / sum(holdings)) %>%
  arrange(desc(pct)) %>%
   slice_head(n = 5)

ggplot(issuer_alloc,
       aes(x = fct_reorder(issuer, pct), y = pct)) +
  geom_col(width = 0.7, fill = corp_blue) +
  add_pct_labels(issuer_alloc, pct) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(issuer_alloc$pct) * 1.15)) +
  labs(title = "Top 5 Issuer (% Nominal)",
       subtitle = "30 Jun 2024 % of portfolio nominal",
       y = "% of nominal") +
  theme_present()

```
----

# Issuer (% mv)

```{r }
# ── 3-C  Issuer allocation (% market value) ─────────────────────────
issuer_alloc <- june %>%
  group_by(issuer,asset_type) %>%         # change if your column differs
  summarise(mv = sum(mv, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = mv / sum(mv)) %>%
  arrange(desc(pct)) %>%
   slice_head(n = 5)

ggplot(issuer_alloc,
       aes(x = fct_reorder(issuer, pct), y = pct)) +
  geom_col(width = 0.7, fill = corp_blue) +
  add_pct_labels(issuer_alloc, pct) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(issuer_alloc$pct) * 1.15)) +
  labs(title = "Top 5 Issuer (% market Value)",
       subtitle = "30 Jun 2024  ·  % of portfolio market value",
       y = "% of mv") +
  theme_present()


```

---

# Top instruments (% Nominal)



```{r }

# ──------------------------Top-5 asset_names (% holdings) ─────────────────────────
top5_instr <- june %>%
  group_by(asset_id, asset_type) %>%      # change if your column differs
  summarise(holdings = sum(holdings, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = holdings / sum(june$holdings, na.rm = TRUE)) %>%
  arrange(desc(holdings)) %>%
  slice_head(n = 5) 
  

ggplot(top5_instr,
       aes(x = fct_reorder(asset_id, pct), y = pct)) +
  geom_col(width = 0.7, fill = corp_blue) +
  add_pct_labels(top5_instr, pct) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(top5_instr$pct) * 1.15)) +
  labs(title = "Top-5 Instruments (% Nominal)",
       subtitle = "30 Jun 2024  ·  % of portfolio nominal",
       y = "% of nominal") +
  theme_present()

```

---

# Top instruments (% MV)


```{r }

# ──------------------------Top-5 asset_names (% mv) ─────────────────────────
top5_instr <- june %>%
  group_by(asset_id, asset_type) %>%      # change if your column differs
  summarise(mv = sum(mv, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = mv/ sum(june$mv, na.rm = TRUE)) %>%
  arrange(desc(mv)) %>%
  slice_head(n = 5) 
  

ggplot(top5_instr,
       aes(x = fct_reorder(asset_id, pct), y = pct)) +
  geom_col(width = 0.7, fill = corp_blue) +
  add_pct_labels(top5_instr, pct) +
  coord_flip(clip = "off") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(top5_instr$pct) * 1.15)) +
  labs(title = "Top-5 Instruments (% Market value)",
       subtitle = "30 Jun 2024  ·  % of portfolio market value",
       y = "% of mv") +
  theme_present()

```

---

# Snapshots changes- Mar24 -> Jun24

```{r asset_trend_holdings }
#---------------------------------------------- ASSET ALLOCATION by Nominal -----------------------------------------------------------------------------------
# allocation by snapshot & asset_type ─────────────
alloc <- snap_clean %>%
  group_by(snapshot, asset_type) %>%
  summarise(holdings_sum = sum(holdings, na.rm = TRUE), .groups = "drop") %>%
  group_by(snapshot) %>%
  mutate(pct = holdings_sum / sum(holdings_sum, na.rm = TRUE)) %>%
  ungroup()

# plot with % labels ────────────────────────────────
pd <- position_dodge(width = 0.75)   # reuse for geom_col and geom_text

ggplot(alloc,
       aes(x = asset_type,
           y = pct,
           fill = factor(snapshot))) +
  geom_col(position = pd, width = 0.75) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            position = pd,
            vjust = -0.4,             # move slightly above bar
            size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(alloc$pct) * 1.15)) +   # add head-room
  labs(title = "Asset Mix (% Nominal)",
       x = "Asset type",
       y = "Portfolio weight (%)",
       fill = "Snapshot") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title    = element_text(hjust = 0.5),   # ← centred
    plot.subtitle = element_text(hjust = 0.5)    # ← centred
  )
```

* *Cash redeployment* of R500m to Swaps(FRNs) and structured (Long Dated SAGBs) as to avoid a cash drag ahead of cutting rate cycle. 
* This seemed like a strategic move to slowly increase duration and benefit from cary through FRNs without exposing the portfolio to excess risk.

---

# Asset Mix (% MV )


```{r asset_trend_mv }
alloc <- snap_clean %>%
  group_by(snapshot, asset_type) %>%
  summarise(mv_sum = sum(mv, na.rm = TRUE), .groups = "drop") %>%
  group_by(snapshot) %>%
  mutate(pct = mv_sum / sum(mv_sum, na.rm = TRUE)) %>%
  ungroup()

# plot with % labels ────────────────────────────────
pd_nominal <- position_dodge(width = 0.75)   # reuse for geom_col and geom_text

ggplot(alloc,
       aes(x = asset_type,
           y = pct,
           fill = factor(snapshot))) +
  geom_col(position = pd, width = 0.75) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            position = pd,
            vjust = -0.4,             # move slightly above bar
            size = 3) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, max(alloc$pct) * 1.15)) +   # add head-room
  labs(title = "Asset Type (% Market Value)",
       #subtitle = "Snapshots: 2024-03-31 vs 2024-06-30",
       x = "Asset type",
       y = "Portfolio weight (%)",
       fill = "Snapshot") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title    = element_text(hjust = 0.5),   # ← centred
    plot.subtitle = element_text(hjust = 0.5)    # ← centred
  )

```

* In march the Ninety one corp bonds was highly concentrated.
* Even though it is still a leading position, however the PM increased risk through structured notes and swaps (SA bank credit and long dated bonds) ahead of decreasing intrest rate colatility in the market.


--- 

# Portfolio Changes Mar24 -> Jun 24

```{r }

# ── 1.  Select March & June, scale billions, compute Δ ─────────────

headline_stat <- snap_clean %>%                  # ← your original aggregation
  group_by(snapshot) %>%
  summarise(
    NAV           = sum(holdings),
    total_mv      = sum(holdings * price),
    port_duration = weighted.mean(active_effective_duration_mac,
                                  w = weight_percent),
    spread_dur    = weighted.mean(active_spread_duration,
                                  w = weight_percent),
    n_positions   = n(),
    .groups = "drop"
  )

tbl_delta <- headline_stat %>% 
  filter(snapshot %in% as.Date(c("2024-03-31", "2024-06-30"))) %>% 
  mutate(across(c(NAV, total_mv), ~ .x / 1e9)) %>%       # R → billions
  mutate(period = if_else(snapshot == as.Date("2024-03-31"),
                          "Mar 2024", "Jun 2024")) %>% 
  select(-snapshot) %>%
  pivot_longer(-period, names_to = "Metric", values_to = "Value") %>% 
  pivot_wider(names_from = period, values_from = Value) %>% 
  mutate(`Δ (Jun–Mar)` = `Jun 2024` - `Mar 2024`) %>% 
  mutate(
    Metric = recode(Metric,
      NAV           = "NAV (R bn)",
      total_mv      = "Total MV (R bn)",
      port_duration = "Eff. MacDur (yrs)",
      spread_dur    = "Spread Dur (yrs)",
      n_positions   = "No. Positions"
    )
  )

# ── 2.  Present with gt ────────────────────────────────────────────
tbl_delta %>% 
  gt(rowname_col = "Metric") %>% 

  ## Header ----
  tab_header(
    title    = md("**Portfolio Headline Metrics**"),
    subtitle = "Change from 31 Mar 2024 to 30 Jun 2024"
  ) %>% 

  ## Number formats ----
  fmt_number(
    columns = c(`Mar 2024`, `Jun 2024`, `Δ (Jun–Mar)`),
    rows    = Metric %in% c("NAV (R bn)", "Total MV (R bn)"),
    decimals = 1
  ) %>% 
  fmt_number(
    columns = c(`Mar 2024`, `Jun 2024`, `Δ (Jun–Mar)`),
    rows    = Metric %in% c("Eff. MacDur (yrs)", "Spread Dur (yrs)"),
    decimals = 2
  ) %>% 
  fmt_integer(
    columns = c(`Mar 2024`, `Jun 2024`, `Δ (Jun–Mar)`),
    rows    = Metric == "No. Positions",
    sep_mark = ","
  ) %>% 

  ## Positive/negative colouring on Δ ----
  tab_style(
    style = cell_text(color = "forestgreen"),
    locations = cells_body(
      columns = `Δ (Jun–Mar)`,
      rows    = `Δ (Jun–Mar)` > 0
    )
  ) %>% 
  tab_style(
    style = cell_text(color = "firebrick"),
    locations = cells_body(
      columns = `Δ (Jun–Mar)`,
      rows    = `Δ (Jun–Mar)` < 0
    )
  ) %>% 

  ## Layout tweaks ----
  cols_align(align = "right") %>% 
  cols_width(everything() ~ px(120)) %>% 
  tab_options(
    table.font.names         = "Verdana, Arial, sans-serif",
    column_labels.font.weight= "bold",
    data_row.padding         = px(4),
    table.border.top.style   = "hidden",
    table.border.bottom.style= "hidden"
  ) %>% 
  tab_style(                                          # zebra stripes
    style = cell_fill(color = "#f7f7f7"),
    locations = cells_body(rows = seq(2, nrow(tbl_delta), 2))
  )
```

* The portfolio Risk metrics have have uptick ahead of the easing rate cycle but were still within mandate limits.

* Duration protetion was trimmed interest rate risk conditions were improving and added FRNs to benefit from carry, increased long duration through structured notes positioning for a long only, as the US CPI has fallen from a high of 9.1 % (Jun-22) to the mid-2 % zone; core-PCE is now just below 2.5%. 

---

# Changes Mar24 -> Jun 24


•	P&L: In nominal terms the portfolio benefited positively from this strategy while managing interest risk at within low levels. Portfolio Value rose to R4,6% from R3.8%.

•	Risk: The Portfolio Effective Duration rose by 16,42% to -0.65yrs from -0.75% in March, with Portfolio Market value. 

* SBK swaps were exited or reached maturity.


```{r, include = FALSE }
#--------------------------------------------------------------- Active Positioning ---------------------------------------------------------------------------------------
# ── 2 · Aggregate active weights ────────────
active_w <- june %>%
   group_by(asset_type) %>% 
  summarise(active_wt = weighted.mean(active_weight_percent,mv, na.rm = TRUE), .groups = "drop")%>%
  arrange(desc(active_wt)) %>%
  mutate(asset_type = factor(asset_type, levels = asset_type))

# ── 3 · Presentation theme with legend & title spacing ─────
theme_present <- function() {
  theme_minimal(base_size = 12, base_family = "Arial") %+replace%
    theme(
      plot.title   = element_text(size = 14, face = "bold",
                                  hjust = 0.5, margin = margin(b = 15)),
      plot.subtitle= element_text(size = 11,  hjust = 0.5,
                                  margin = margin(b = 6)),
      axis.title.y = element_blank(),
      axis.title.x = element_text(margin = margin(t = 6)),
      axis.text    = element_text(color = "grey30"),
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "top",
      legend.title       = element_blank(),
      plot.margin        = margin(10, 25, 10, 10)   # TLBR
    )
}

# Corporate colours
over_col <- "#2C7FB8"   # Overweight  – blue
under_col <- "#D7191C"  # Underweight – red

# ── 4 · Plot: active weight vs benchmark ───────────────────
ggplot(active_w,
       aes(x = asset_type,
           y = active_wt,
           fill = active_wt > 0)) +               # TRUE = overweight
  geom_col(width = 0.7) +
  geom_text(aes(label = percent(active_wt, accuracy = 0.1)),
            hjust = ifelse(active_w$active_wt >= 0, -0.15, 1.15),
            size  = 3.1, colour = "black") +
  coord_flip(clip = "off") +
  scale_fill_manual(values = c(`TRUE` = over_col,
                               `FALSE` = under_col),
                    labels  = c(`TRUE` = "Overweight",
                                `FALSE`= "Underweight")) +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits  = c(min(active_w$active_wt) * 1.3,
                                 max(active_w$active_wt) * 1.3)) +
  labs(title    = "Asset Type Active- Jun 2024",
      
       y = "Active weight (%)") +
  geom_hline(yintercept = 0, colour = "grey50") +
  theme_present()


```

---

# Risk Evolution (%)

```{r }

# ── 1.  Declare the metrics you want to show ───────────────────────────
metric_labels <- c(
beta_p         = "Beta (DV01 ratio)",
 spread_duration_active    = "Active Spread duration (yrs)",
  credit_spread_dur_active  = "Active Credit spread-dur (yrs)",
  tracking_error_ex_ante = "Tracking-error (%)"
)

# ── 2.  Reshape risk_ts to long format ────────────────────────────────
plot_data <- risk_ts %>%                         # <–– your time-series tibble
  select(reference_date, all_of(names(metric_labels))) %>%
  pivot_longer(-reference_date, names_to = "metric", values_to = "value") %>%
  mutate(metric = metric_labels[metric])

# ── 3.  Plot: small multiples (2×2) ────────────────────────────────────
p <- ggplot(plot_data, aes(reference_date, value)) +
  geom_line(color = "#1f77b4", linewidth = 0.9, na.rm = TRUE) +
  geom_point(color = "#1f77b4", size = 1.4, na.rm = TRUE) +
  facet_wrap(~ metric, scales = "free_y", ncol = 2) +
  scale_x_date(
    date_breaks  = "3 month",
    date_labels  = "%b %Y",
    expand       = expansion(mult = c(0.01, 0.02))
  ) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    title    = "Portfolio risk metrics over time",
    #subtitle = "Beta, active duration, spread-duration and tracking-error",
    x        = NULL,
    y        = NULL,
    
    caption  = "Source: internal risk engine · Data as at latest valuation"
  ) +
  theme_minimal(base_size = 12, base_family = "Helvetica") +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(margin = margin(b = 10)),
    panel.spacing   = unit(1.3, "lines"),
    axis.text.x     = element_text(angle = 45, hjust = 1),
    strip.text      = element_text(face = "bold")
  )

print(p)
```

* Tracking Error > 1 % (below 1.5 reg limit): shows conviction. The managers aren’t hugging ALBI—they’re using just under 80 % of their risk budget to pursue active returns. Yet still within risk limits of 1.5%.

* Beta > 1: the portfolio has a residual long-duration bias relative to benchmark. Managers are showing that the managers have entered duration bias view taking more slight duration thatn the benchmark.

* Spread duration exposure increased with active credit spread DV01 at 1.4 yrs, the fund will gain if credit spreads tighten—but lose if SA credit stress returns.
* Between 15th to 28th of May TE Rose rose from ~0.84 to 1.14%? This seems like a deliberate re-risking with plan to benefit from the rate cut cycle especially from long dated bonds.

---

# Data anomalities spotted & Fixes

* Prices < 0 for the asset swaps which inflated mac durations to >1000- Seen mainly from ASW FRNs(These are real exposures)
* Put temp placeholders- approx prices at par(100) as not tonskew risk numbers while waiting for ops to investigate and update
* Cash and Ninety one zero coupon bond prices were missing - approximate cash with 1 and Ninety Corp with par 100
* Negative holdings for Repo - immaterial to the overall portfolio~ Included but put zero normalise the holdings.
* The 1 day spike in TE on 28th May worth Investigating
* Automated a report for all these data anomalities(run the report it would automatically output to the same folder this code is saved)

